---
title: "bikeshare-report"
author: "Karl Nagy"
date: "01/24/2021"
output: html_document
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE,error = FALSE, echo = FALSE) 
```


# Key Questions
* What differentiates annual members and casual riders?
* How can casual riders be converted to annual members to drive future growth?

# Summary
Casuals and members differ in several ways  

* members ride during weekdays more than casuals
* member riding behavior is less seasonal than casuals
* members and casuals bike similar distances
* members ride faster than casuals

These differences in behavior have implications for a marketing campaign attempting to convert casuals to members. The guiding philosophy of such a campaign will be to identify member-like behavior in casuals, and appeal to them for membership.

# Data Sources and Security

Historical bikeshare use [dataset](https://divvy-tripdata.s3.amazonaws.com/index.html) obtained from *Divvy*, a Chicago-based bikeshare company  

Google drive used for storage of all files




# Data Cleaning and Manipulation
Initial cleaning and manipulation was done in BigQuery SQL editor, then transferred to R.




## BigQuery SQL Data Cleaning

###  Added
* `day_of_week` column created from `started_at`, with integers 1-7 corresponding to Sunday-Saturday

###  Changed
* 2020.10 and 2020.11 tables use integer values for `start_station_id` and `end_station_id` columns, changed to string values to match the other tables

###  Fixed
* removed rows where `started_at` occurred after `ended_at`



**End product:** bikeshare_v4.csv




## Data Cleaning and Wrangling in R

### Packages Used
The following were used when in R
```{r}
install.packages("tidyverse", repos='http://cran.us.r-project.org')
library(tidyverse)
library(purrr)
library(lubridate)
library(here)
library(hms)
library(scales)
library(rstudioapi)
install.packages("ggmap", repos='http://cran.us.r-project.org')
library(ggmap)
install.packages("pwr", repos='http://cran.us.r-project.org')
library(pwr)
install.packages("hrbrthemes", repos='http://cran.us.r-project.org')
library(hrbrthemes)
install.packages("googledrive", repos='http://cran.us.r-project.org')
library(googledrive)
install.packages("kableExtra", repos='http://cran.us.r-project.org')
library(kableExtra)
```
### bikeshare_v5
* Added: `ride_length` column created from `ended_at` - `started_at` columns

### bikeshare_v6
* Changed: replaced NA values of start and end station names and IDs with "undocked"

### bikeshare_v7
* Changed: `rideable_type` instances of "docked bike" to "classic_bike"

### bikeshare_v8
* Changed: converted `ride_length` to a `lubridate` duration object for easier reading

### bikeshare_v9
* Added: created `ride_length_min`, ride_length expressed in minutes as a numeric datatype
* Changed: `ride_length` to `ride_length_dur` for clarity

### bikeshare_v10
* Fixed: removed outlier rows with a `ride_length_dur` longer than 1440 minutes(1 day) or shorter than 30 seconds

### bikeshare_v11
* Added: `area` column with "Hyde_Park", "Evanston", "Z1", "Z2", and "noncluster" categories by setting boundaries of  `start_lat` and `start_lng` values, based on [Divvy's Zone classification](https://account.divvybikes.com/map) and interpreted below

   Z1:  start_lat > 41.822978 & start_lat < 42.0219536 & start_lng > -87.690118 & start_lng < -87.593127
   
   Z2_nocollege:   start_lat > 41.643837 & start_lat < 42.0219536 & start_lng > -87.836541 & start_lng < -87.524491 & isFALSE(start_lat > 41.77 & start_lat < 41.82 & start_lng > -87.61 & start_lng < -87.56)
   
   Hyde Park: start_lat > 41.77 & start_lat < 41.82 & start_lng > -87.61 & start_lng < -87.56
   
   Evanston:  start_lat > 42.0219536 & start_lat < 42.07 & start_lng > -87.69 & start_lng < -87.65
                  


### bikeshare_v12
* Removed: examined "noncluster" rows and found them to be 44 rows outside of Zone 2 boundaries

### bikeshare_v13
* Fixed: removed rows less than 1 minute, in keeping with Divvy's policy on their data

### bikeshare_v14
* Added: calculated the Manhattan distance between start and end lat-lng data columns, stored in miles as `ride_ml`
* Added: calculated average speed of ride  from `ride_length_min` and `ride_ml`, stored in mph as `speed_mph`

### bikeshare_v15
* Changed: `area` column parameters altered to distinguish Evanston from Northwestern University

 * Z1: start_lat > 41.822978 & start_lat < 42.0219536 & start_lng > -87.690118 & start_lng < -87.593127
   
 * Hyde_Park: start_lat > 41.77 & start_lat < 41.82 & start_lng > -87.61 & start_lng < -87.56
   
 * Z2_nocollege: start_lat > 41.643837 & start_lat < 42.0219536 & start_lng > -87.836541 & start_lng < -87.524491
   
 * E_NWU: start_lat > 42.044047 & start_lng > -87.684714 & start_lat < 42.07 & start_lng < -87.65
   
 * E_nocollege: start_lat > 42.0219536
                 
    The E_nocollege includes a section that technically is part of Z1, a 2x4 block section just south of Cavalry Catholic Cemetery

### bikeshare_v16
* Changed: `E_nocollege` > `Ev_nocollege`
           `E_NWU` > `Ev_NW_College`
* Fixed: readjusted Hyde_Park boundaries to specific streets
  * `Hyde_Park`    - E. 63rd St. in the South - (41.780428 lat)
                   - S. Martin Luther King Dr. in the West - (-87.615707 lng)
                   - E. Hyde Park Blvd in the North - (41.802297 lat)
                   - S. Stony Is Ave. in the East - (-87.586710 lng)
                 
  * `Ev_NW_College`- Bordered by Lake St. in the South - (42.044047 lat)
                   - Maple Ave. in the West - (-87.684714 lng)
               
 * `Ev_nocollege`  - Bordered by W. Howard St. in the South - (42.0219536 lat)
                   - excluding the `Ev_NW_College` area
                   
 * `Z1`            - Bordered by W. Pershing Rd in the South - (41.822978 lat)
                   - N Western Ave. in the West - (-87.690118 lng)
                   - W Howard St. in the North - (42.0219536 lat)
                   - these borders omit a small chunk of the north of `Z1`, 
                      assigning it instead to `Ev_nocollege`
                      
 * `Z2_nocollege`  - Out to Indiana state line in the East - (-87.524491 lng)
                   - W Howard St. in the North - (42.0219536 lat)
                   - excludes `Z1` and `Hyde_Park`

### bikeshare_v17
* Removed: `ride_length_dur` was no longer needed 
```{r}

drive_download(
  as_id("1ZiEa7xBAaW0IXlQh97FCYMYOdkRuHL6u"),overwrite = TRUE) ## requires google account for use
bikeshare_v17 <- read_csv("bikeshare_v17.csv", 
    col_types = cols(
     ride_id = col_character(),
     member_casual = col_character(),
     rideable_type = col_character(),
     started_at = col_datetime(),
     ended_at = col_datetime(),
     day_of_week = col_double(),
     start_station_name = col_character(),
     start_station_id = col_character(),
     end_station_name = col_character(),
     end_station_id = col_character(),
     start_lat = col_double(),
     start_lng = col_double(),
     end_lat = col_double(),
     end_lng = col_double(),
     ride_length_min = col_double(),
     area = col_character(),
     ride_ml = col_double(),
     speed_mph = col_double()))
```




### Weekday Bike Use
```{r}
bikeshare_v17 %>%
  mutate(day_name = case_when(day_of_week == 1 ~ "Su",
                              day_of_week == 2 ~ "M",
                              day_of_week == 3 ~ "T",
                              day_of_week == 4 ~ "W",
                              day_of_week == 5 ~ "Th",
                              day_of_week == 6 ~ "F",
                              day_of_week == 7 ~ "S")) %>%
  mutate(day_name_2 = fct_relevel(day_name,"M","T","W","Th","F","S","Su")) %>%
  group_by(day_name_2,member_casual) %>%
  summarise(count = n()) %>%
  
ggplot(aes(x=day_name_2,
           y = count,
           group = member_casual,
           color = member_casual)) +
  geom_point(position = "dodge",
             size = 2,
             show.legend = FALSE) +
  geom_line(position = "dodge",
            size = 1,
            show.legend = FALSE) +
  scale_y_continuous("Total Rides",
                     labels = comma,
                     limits = c(0,700000),
                     expand = c(0,0)) +
  annotate("text",x = 2.5, y = 450000, label = "Members") + 
  annotate("text",x = 2.5, y = 200000, label = "Casuals") +
  labs(title = "Casuals like weekends",
       subtitle = "Members ride weekdays",
       x = NULL,
       caption = "Data: Oct-2020 to Oct-2021") +
  theme_ipsum()


```
Members use bikes more during the weekday than casuals, who prefer the weekend. This is likely due to members using the bikes for work/school, and casuals for pleasure. The trend is especially prominent in the 'Hyde_park` cluster around the university campus, where students have reason not only for home-work-home transit, but multiple short trips from class to class, nearby restaurants, and study locations. It is worth mentioning that without individual data to confirm how many trips per person are taken per day or more detailed individual information to confirm a lifestyle pattern, this hypothesis can only go so far.




### Seasonal Bike Use
```{r}
bikeshare_v17 %>%
        
ggplot(aes(x = started_at,
           fill = member_casual)) +
  geom_density(alpha = .4) +
  scale_y_continuous("Rider Density",
                     labels = NULL, breaks = NULL) +
  scale_x_datetime(date_breaks = "months", date_labels = "%b",
                   minor_breaks = NULL) +
  scale_fill_discrete(name = "") +
  labs(title = "Yearly Rider Density",
       subtitle = "Comparing Members and Casuals",
       x = element_blank(),
       y = element_blank(),
       caption = "Data: Oct-2020 to Oct-2021") +
  theme_ipsum()
```

Chicago winters are harsh, and total rides across both members and casuals show this; Rides decrease outside of temperate months and increase when weather is pleasant. Anecdotal observation of recorded temperatures show that total rides drop harshly when the maximum temperature reaches freezing or below, likely the bitter cold makes biking less attractive. During Jan. 26 - Feb. 15 for example, Chicago had a record 34 in. of snow, which corresponds to the lowest point of ride density.

Though both members and casuals are affected by the seasons, members have less of a ride *increase* than casuals in temperate months, and less of a ride *decrease* in colder months. This more level yearly behavior taken with the weekday use pattern seems to suggest a clear theme: members ride for function, casuals ride for pleasure, and bike riding as a whole is seasonal.




### Distance, speed, and ride time
```{r}
fun_peak <- function(x) { ## this function finds the corresponding x value for a max y value of the 
  d <- density(x)         ## density of a data frame
  d$x[which.max(d$y)]
  }


bikeshare_v17 %>%
  filter(ride_length_min < 46.618) %>% ## 92.6% of all rows, outliers eliminated through IQR method
  
ggplot(aes(x = ride_length_min,
           fill = member_casual)) +
  geom_density(alpha = .4) +
  expand_limits(y = 0.1) +
  scale_y_continuous("Rider Density",
                     labels = NULL, breaks = NULL) +
  scale_fill_discrete(name = "") +
  labs(title = "Ride Time",
       subtitle = "Members take shorter rides",
       x = "Minutes",
       y = element_blank(),
       caption = "Data: Oct-2020 to Oct-2021") +
  theme_ipsum()

bikeshare_v17 %>%
  filter(ride_length_min < 46.618,
         ride_ml != 0) %>% ## 92.6% of all rows, outliers eliminated through IQR method
  
ggplot(aes(x = speed_mph,
           fill = member_casual)) +
  geom_density(alpha = .4) +
  expand_limits(y = 0.2) +
  scale_y_continuous("Rider Density",
                     labels = NULL, 
                     breaks = NULL) +
  scale_x_continuous(limits = c(0,20)) +
  scale_fill_discrete(name = "") +
  labs(title = "Average Speed",
       subtitle = "Members go faster than Casuals",
       x = "MpH",
       y = element_blank(),
       caption = "Data: Oct-2020 to Oct-2021") +
  theme_ipsum()

bikeshare_v17 %>%
  filter(ride_length_min < 46.618,
         ride_ml != 0) %>% ## 92.6% of all rows, outliers eliminated through IQR method
  
ggplot(aes(x = ride_ml,
           fill = member_casual)) +
  geom_density(adjust = 4,
               alpha = .4) +
  expand_limits(y = 0.2) +
  scale_y_continuous("Rider Density",
                     labels = NULL, 
                     breaks = NULL) +
  scale_x_continuous(limits = c(0,8)) +
  scale_fill_discrete(name = "") +
  labs(title = "Distance",
       subtitle = "Members and Casuals ride a Similar Distance",
       x = "Miles",
       y = element_blank(),
       caption = "Data: Oct-2020 to Oct-2021") +
  theme_ipsum()




```
Members and Casuals tend to cover a similar distance, though Members tend to shorten their ride by going faster than Casuals.

Considering that Chicago bikeshare rides compete with walking, electric scooters, public transportation, and automobile use, longer or shorter distances further from the *peak of 0.75 to 1.5mi.* may cause potential riders to use an alternative means of transport instead. 




### Rideable type use
```{r}
bikeshare_v17 %>%                                      
  filter(ride_length_min < 46.4) %>%
  group_by(member_casual) %>%
  summarise(ebike_rate = sum(rideable_type == "electric_bike")/n(),
            dist = mean(ride_ml),
            speed = mean(speed_mph),
            count = n()) %>%
  
ggplot(aes(x = member_casual,
           y = ebike_rate,
           fill = member_casual)) +
  geom_col(position = "dodge",
           width = .3) +
  geom_text(aes(label = percent(ebike_rate,1)),
            vjust = -.75) +
  scale_y_percent("Percent of rides taken on Ebikes",
                  limits = c(0,1)) +
  scale_x_discrete("",
                   breaks = NULL) +
  scale_fill_discrete(name = "") +
  labs(title = "Members vs. Casuals on ebikes",
       subtitle = "Casuals have only a slight preference",
       caption = "Data: Oct-2020 to Oct-2021") +
  theme_ipsum()


bikeshare_v17 %>%                                     
  filter(ride_length_min < 46.4,
         ride_ml != 0) %>%
  mutate(area_grouped = case_when(area == "Ev_nocollege" ~ "Z1",   ## Z1 and Z2 together to check price
                                  area == "Z1" ~ "Z1",
                                  area == "Ev_NW_College" ~ "Z1",
                                  area == "Z2_nocollege" ~ "Z2",
                                  area == "Hyde_Park" ~ "Z2")) %>%
  group_by(area_grouped,member_casual) %>%
  summarise(ebike_rate = sum(rideable_type == "electric_bike")/n(),
            dist = mean(ride_ml),
            speed = mean(speed_mph),
            count = n()) %>%
  
ggplot(aes(x = area_grouped,
           y = ebike_rate)) +
  geom_col(aes(fill = member_casual),
           position = "dodge",
           width = .5) +
  geom_text(aes(label = percent(ebike_rate,1),
             group = member_casual),
             position = position_dodge(width = .5),
            vjust = -.75) +
  scale_y_percent("Percent of rides taken on Ebikes",
                  limits = c(0,1)) +
  scale_x_discrete("") +
  scale_fill_discrete(name = "") +
  labs(title = "Zone 2 loves E-bikes",
       subtitle = "Zone has a much larger effect than membership",
       caption = "Data: Oct-2020 to Oct-2021") +
  theme_ipsum() 

```
E-bike vs. classic bike use is not about the absolute cost, but the relative cost of using the other mode of transport. 

Consider the example the cost of a six minute e-bike ride in each zone, with the bike being returned to a dock, for a member or a casual, contrasted with the classic bike alternative cost:

```{r}

rideable <- c("E-bike","Classic")
Z1_Casual <- c("$4.50","$3.30")
Z1_Member <- c("$0.90","$0.00")
Z2_Casual <- c("$3.30","$3.30")
Z2_Member <- c("$0.00","$0.00")

comparison <- data.frame(rideable,Z1_Casual,Z1_Member,Z2_Casual,Z2_Member)

comparison %>%
  select(rideable,everything()) %>%
  arrange(desc(rideable)) %>%
  kable(col.names = c("","Casual","Member","Casual","Member")) %>%
  column_spec(c(1:5), border_right = T) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  collapse_rows(columns = 1) %>%
    add_header_above(c(" " = 1, "Z1" = 2, "Z2" = 2))
```

Z2 riders have no cost difference between an electric or classic bike, and so use the higher end option of e-bikes more frequently. Z1 riders may decide to take a bike ride, and having locked in that decision now consider cost, find e-bikes too expensive, and take a classic bike instead.

Z1 members receive a strong discount on e-bike use when compared with Z1 casuals, but still use e-bikes less. This could be that members take more bike rides per-week, and thus the yearly cost of e-bike use for an individual is higher for members than casuals. For verification of this hypothesis, data on how often a given individual takes rides is needed. 

People who rent e-bikes ride 18% further than those who do not, but ride distance is a much smaller factor on e-bike use than the relative cost of e-bike vs. classic.




## Frequent Riders
Membership is a bulk ride discount. In theory, people who ride the most would see the most cost savings benefit. Without individualized data access there is no supporting evidence for this, but the underlying logic is so fundamental that it merits discussion.

Casuals who pass the breakeven point for ride savings of more than two 30-minute rides per month stand to save money from a membership, assuming they keep this behavior up throughout the year. Not purchasing a membership when they could save money may indicate a reluctance to commit to a lifestyle, therefore strategies to appeal to this group should focus on making commitment easier.




## Limitations/Further information needs

### Individual rider data
Since membership is a flat yearly fee for rides below a certain duration, its main pitch to casuals is that of bulk discount: more rides for less money. Even without individual data, this alone means casuals who ride multiple days per week, or multiple times per day are prime targets for advertising. Being able to compare individuals would allow the "members have a commute" hypothesis to be tested.

Other information about individuals like occupation, age, residence, yearly earnings, and transportation mode habits could be used to better profile customer segments and identify patterns and needs. One of the key issues for converting casuals to members is knowing if a casual is a resident who could use a membership, or an out-of-towner who would not buy a membership, no matter their casual use patterns. Separating resident casuals from nonresident casuals would be key in any targeted marketing effort, and without individualized data this cannot be done.

### Individual bike data
Ebikes may be so popular that their actual use is limited by the number of them in circulation. With individual bicycle data, the uptime per bike could be found, and a better sense of e-bike popularity established beyond current data. Knowing other information like locked vs. docked e-bike use could help identify whether docking availability may act as a confounding variable. Without this crucial information, no conclusion can be made as to member/casual e-bike preference.




## Recommendations
The goal is to convert Casual riders to Members. With this in mind, an ideal ad campaign would find casuals with member-like behaviors and appeal to them. Here are four primary behavior types to look for in casuals, and strategies for each:

### 1) Weekday riders
* Offer a free week of membership to foster a lifestyle switch

### 2) Fast riders
* Highlight rides as a fast, self-paced commute alternative
* Contrast with other transport as being greener, traffic-free, and self-controlled

### 3) Warm season riders
* Offer limited-time deals on an annual membership in the June-September time frame, when casuals see the best ride weather quality
* Highlight good weather in ads

### 4) Frequent riders 
* Offer free e-bike usage for a certain number of miles biked as part of a member rewards program to induce commitment
